A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-18 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewed ensemble space

We start by assuming that physical laws deal with experimentally reproducible relationships, which means that they must describe relationships between statistical ensembles. Conceptually, a statistical ensemble is the collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one). An ensemble space is the collection of all possible ensembles defined in a physical theory, together with a basic mathematical structure that all physical theories must possess.

An ensemble space $\Ens$ will need to satisfy a three basic requirements. First, it needs to guarantee experimental verifiability. Mathematically, the ensemble space must be a $\mathsf{T}_0$ second countable topological space, where each open set represents an experimentally verifiable statement.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Second, it needs to allow statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$. Mathematically, this constrains the ensemble space to be a convex set.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i}^{\infty} p_i \ens_i$ converges in the space or not.} Third, an entropy function must be defined that characterizes the variability of the elements within an ensemble.

Additionally, we will require that ensembles can be identified through real valued statistical quantities. \footnote{It is an open question whether all ensemble spaces must possess this property. In a previous work we showed that the dense linear ordering require highly idealized assumptions.} A statistical quantity represents the expectation of a real valued quantity, and is therefore an affine continuous functions $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$. The last requirement is that there exists a family $F_i$ of statistical quantities that fully identify each ensemble. That is, $F_i(p \ens[a] + \bar{p} \ens[b]) = p F_i(\ens[a]) + \bar{p} F_i(\ens[b])$ and $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$.

These requirements interact with each other leading to a rich mathematical structure that connect to many of the structures we use in mathematical physics. In this paper, we will examine only connections to measure theoretic representations.

\textbf{Ensemble spaces as convex sets.} In this work, an ensemble space $\Ens$ is a convex subset of a metrizable second countable locally convex topological vector space $V$.\footnote{The entropy constraints forces the topology to be $\mathsf{T}_1$ the convex space to embed into a vector space. A second countable $\mathsf{T}_1$ TVS is metrizable. The family of statistical quantities induce a family of semi-norms on the space, which must be countable due to second countability, which makes the TVS locally convex.} The entropy is function $S : \Ens \to \mathbb{R}$ that satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.}  Intuitively, the variability (i.e. the entropy) cannot decrease during mixing, and will have an upper bound when mixing two ensembles whose instances are always distinguishable.

\textbf{Orthogonality and separatedness.} We will say that two ensembles are orthogonal, noted $\ens[a] \ortho \ens[b]$, when their mixture gives a maximal entropy increase. This will recover the notion of orthogonality for both classical and quantum spaces. Another notion of distinction between two ensembles is in terms of the mixtures: two ensembles are separate, noted $\ens[a] \separate \ens[b]$, they cannot be expressed as a mixture of a common ensemble. That is, there is no $\ens[c]$ such that $\ens[a] = \alpha \ens[c] + \bar{\alpha} \ens[d]$ and $\ens[b] = \beta \ens[c] + \bar{\beta} \ens$ from some $\alpha, \beta \in (0,1]$ and $\ens[d],\ens \in \Ens$. In classical spaces, these two notions coincide: two ensembles are orthgonal and separate if and only if the support of the respective probability distributions overlap on a non-measure-zero set. In quantum mechanics, these two notions are different: any two pure states are separate but are not orthogonal.

To summarize, a repeatedly verifiable physical theory requires ensembles, which must be defined experimentally, allow statistical mixing and provide an entropy. Additionally, we require the existence of enough statistical quantities to identify each ensemble. Under these assumptions, an ensemble space $\Ens$ is a convex set of a metrizable second countable locally convex topological vector space $V$. We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.

>>>>

We start by assuming that physical laws deal with experimentally reproducible relationships, which implicitly requires the notion of statistical ensembles. Conceptually, a statistical ensemble is the collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one). A physical law, then, ultimately expresses relationship among ensembles. A minimal requirement for a physical theory, then, is to provide an ensemble space: the collection of all possible ensembles within the theory, together with a basic mathematical structure that all physical theories must possess.

An ensemble space $\Ens$ will need to satisfy three basic requirements. First, it needs to guarantee experimental verifiability. Mathematically, the ensemble space must be a $\mathsf{T}_0$ second countable topological space, where each open set represents an experimentally verifiable statement.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Second, it needs to allow statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$. Mathematically, this constrains the ensemble space to be a convex set.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i}^{\infty} p_i \ens_i$ converges in the space or not.} Third, an entropy function must be defined that characterizes the variability of the elements within an ensemble.

Additionally, we will require that ensembles can be identified through real valued statistical quantities.\footnote{It is an open question whether this is an additional assumption or it can be derived from the other requirements.} A statistical quantity represents the expectation of a real valued quantity, and is therefore an affine continuous functions $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$. The last requirement is that there exists a family $F_i$ of statistical quantities that fully identify each ensemble. That is, $F_i(p \ens[a] + \bar{p} \ens[b]) = p F_i(\ens[a]) + \bar{p} F_i(\ens[b])$ and $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$.\footnote{For example, on the continuum this means that the moment problem is solvable: you can identify the distribution by the expectation of the polynomials of position and momentum. Similarly, in quantum mechanics means that the state can be reconstructed through quantum tomography.}

Each physical theory satisfies these axioms, even though it may be represented very different mathematically. Classical mechanics, for example, is typically described in terms of the geometry of phase space. Quantum mechanics is most often defined in terms of complex inner product spaces, but can be reframed in terms of quasi-probability distributions. Therefore, while the above requirements are needed for a physical theory, it does not necessarily mean that it is convenient to describe problems or make calculations in terms of those definitions. In fact, those requirements interact with each other leading to a rich mathematical structure that connect to many of the structures we use in mathematical physics, and its goal is mainly to see how the different representations are connected.

The goal of this paper is to examine the different way that measure theory can be used to represent ensembles in any given physical theory.

\textbf{Ensemble spaces as convex sets.} For the purpose of measure theoretic representations, it is best to understand an ensemble space $\Ens$ as a convex subset of a metrizable second countable locally convex topological vector space $V$.\footnote{The entropy constraints forces the topology to be $\mathsf{T}_1$ the convex space to embed into a vector space. A second countable $\mathsf{T}_1$ TVS is metrizable. The family of statistical quantities induce a family of semi-norms on the space, which must be countable due to second countability, which makes the TVS locally convex.} The entropy is a function $S : \Ens \to \mathbb{R}$ that satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.}  Intuitively, the variability (i.e. the entropy) cannot decrease during mixing, and will have an upper bound when mixing two ensembles whose instances are always distinguishable.

Using the convex structure, we say that two ensembles are separate, noted $\ens[a] \separate \ens[b]$, if they cannot be expressed as a mixture of a common ensemble. That is, there is no $\ens[c]$ such that $\ens[a] = \alpha \ens[c] + \bar{\alpha} \ens[d]$ and $\ens[b] = \beta \ens[c] + \bar{\beta} \ens$ from some $\alpha, \beta \in (0,1]$ and $\ens[d],\ens \in \Ens$. No ensemble is separate from itself. An internal point is an ensemble that is not separate from any ensemble (i.e. it is the mixture of any ensemble). An extreme point is an ensemble that is separate from all others (i.e. it is only a mixture of itself).

TODO: figure with classical triangle and quantum Bloch Ball

The set of all extreme points $X$, as we will see, plays a major role. However, on physical grounds, these may not be proper ensembles. In classical mechanics, these would corresponds to Dirac measure: probability distributions all concentrated at a single point of phase space. Since we cannot reliably prepare a particle with infinite precision, these are not ensemble and they are not part of $\Ens$. For the purpose of Choquet theory, we will assume that there is a way to find the closure $E \supset \Ens$ that includes all extreme points $X$. Moreover, we will also assume $E$ to be compact, even though it will not be. How to properly make these identification, and how to go beyond the compact case, is outside of the scope of this particular article.\footnote{An early attempt used the notion of orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. This does not work. An approach, currently under investigation, is to study limits over statistical quantities. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} Here we will use $\Ens$ to indicate the general case and $E$ to indicate this ``compact closure.''

\textbf{Orthogonality and separatedness.} We will say that two ensembles are orthogonal, noted $\ens[a] \ortho \ens[b]$, when their mixture gives a maximal entropy increase. This will recover the notion of orthogonality for both classical and quantum spaces. Conceptually, these corresponds to two ensembles where all instances are distinct from the other: they are mutually exclusive and a single instance is enough to distinguish between the two cases.

In classical spaces, orthogonality and separatedness coincide: two ensembles are orthgonal and separate if and only if the support of the respective probability distributions overlap on a non-measure-zero set. In quantum mechanics, these two notions are different: any two pure states are separate but are not orthogonal. For example, pure ensembles of spin up and spin down are orthogonal, mutually exclusive, precisely because it takes one instance to determine which one is which. Pure ensembles of spin up and spin right, however, are not orthogonal precisely because, in some cases, their instance cannot be perfectly discriminated. They are, however, orthogonal, because they are both pure states, one is not the mixture of other ensembles: they are extreme points.

It is the ease with which such distinctions between the classical and quantum case can be made that is the main appeal for this structure. We will see over and over how mutual exclusivity is linked to classical probability and its failure in quantum, and other non-classical, theories.

To summarize, a repeatedly verifiable physical theory requires ensembles, which must be defined experimentally, allow statistical mixing and provide an entropy. Additionally, we require the existence of enough statistical quantities to identify each ensemble. Under these assumptions, an ensemble space $\Ens$ is a convex set of a metrizable second countable locally convex topological vector space $V$. In some cases, for simplicity, we will require $\Ens$ to admit a ``compact closure'' $E$, so that it includes the set $X$ of its extreme points. We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.
